---
title: "Usage of the hFDR package with Gaussian graphical model"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GaussianGraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---


This vignette illustrates the usage of the `hFDR` package for estimating the FDR of variable selection procedures in the Gaussian graphical model.

For simplicity, we will use synthetic data constructed from multivariate Gaussian distribution with a sparse precision matrix.


```{r, results='hide', message=FALSE, warning=FALSE}
set.seed(2024)
library(hFDR)
```

```{r define-problem, results='hide', message=FALSE, warning=FALSE}
# Problem parameters
p <- 15            # number of variables
n <- 500           # number of observations
k <- 10            # number of variables with nonzero coefficients
amplitude <- 0.2   # signal amplitude

# Generate a sparse precision matrix (positive definte)
nonzero.row <- sample(p-1, k)
nonzero.col <- sapply(nonzero.row, function(row){sample((row+1):p, 1)})
nonzero <- sapply(1:k, function(ind){hFDR:::pair_to_index(nonzero.row[ind], nonzero.col[ind])})

Precision <- diag(p)
for(ind in 1:k) Precision[nonzero.row[ind], nonzero.col[ind]] <- amplitude
Precision <- (Precision + t(Precision))/2

# Generate the variables from a multivariate normal distribution
Sigma <- solve(Precision)
X <- matrix(rnorm(n*p), n) %*% chol(Sigma)

# generate lambda sequence for lasso
n_lambda <- 40
lambda_max <- max(abs(t(X) %*% X)) / n
lambda_min <- lambda_max / 100
lambda <- lambda_max * (lambda_min/lambda_max)^((0:n_lambda)/n_lambda)
```

Example 1. Graphical Lasso selection for nonzero entries in the precision matrix
--------------

To start with, we call `hFDR` for graphical lasso with the default settings.
```{r hFDR-glasso, message=F, warning=F, fig.width=7, fig.height=4.5}
hFDR.res <- hFDR(X, y, model = "gaussgraph", select = select.glasso, lambda = lambda,
                 psi = "pval", n_cores = 2)
plot(hFDR.res)
```

Now estimate standard error of the FDR estimator and plot it in company with 
cross-validation.
```{r hFDR-glasso-cv, message=F, warning=F, fig.width=7, fig.height=4.5}
# do cross-validation for glasso
cv.res <- cv.gaussgraph(X, lambda, precision.est.glasso, select = select.glasso,
                        type.measure = "-loglikelihood", nfold = 10)
hFDR.res <- hFDR(X, y, model = "gaussgraph", select = select.glasso, lambda = lambda,
                 psi = "pval", se = T,
                 n_sample.hfdr = 20, n_sample.se = 10, n_cores = 2)
plot(hFDR.res, cv.res)
```

Compare with FDR and FDP in the simulation setting.
```{r hFDR-FDR-lasso, message=F, warning=F, fig.width=7, fig.height=4.5}
calc_FDP <- function(selected, nonzero){
  p <- NROW(selected)
  nonzero <- (1:p) %in% nonzero
  colSums(selected * !nonzero) / pmax(colSums(selected), 1)
}

FDP <- calc_FDP(select.glasso(X, lambda), nonzero)
FDR <- sapply(1:100, function(mc_i){
  X.mc <- matrix(rnorm(n*p), n) %*% chol(Sigma)
  calc_FDP(select.glasso(X.mc, lambda), nonzero)
})
FDR <- rowMeans(FDR)

plot(hFDR.res, cv.res)
lines(x = -log(lambda), y = FDP, col = "black", lty = 2)  # dashed black line for FDP
lines(x = -log(lambda), y = FDR, col = "black", lty = 1)  # solid black line for FDR
```

Example 2. User-defined selection procedures with Monte-Carlo calculation
--------------
The selection procedure is up to the user to choose. Here give an example with selection is via p-value thresholding.
```{r hFDR-user-defined, message=F, warning=F, fig.width=7, fig.height=4.5}
# select by p-value thresholding
select.my <- function(X, lambda){
  sapply(lambda, function(t){!psi.gaussgraph(X, 1:(p*(p-1)/2), threshold = t)$psi})
}
# self-defined psi function -- using 0.5 p-value thresholds
psi.my <- function(X, pair_inds){
  psi.gaussgraph(X, pair_inds, threshold = 0.5)
}

lambda <- seq(0, 0.5, length.out = 10)

hFDR.res <- hFDR(X, y, model = "gaussgraph", select = select.my, lambda = lambda,
                 psi = psi.my, se = F, n_sample.hfdr = 20, n_cores = 2)

plot(hFDR.res, sign.lambda = 1, log_lambda = F)

# compare with FDR and FDP in the simulation setting
FDP <- calc_FDP(select.my(X, lambda), nonzero)
FDR <- sapply(1:100, function(mc_i){
  X.mc <- matrix(rnorm(n*p), n) %*% chol(Sigma)
  calc_FDP(select.my(X.mc, lambda), nonzero)
})
FDR <- rowMeans(FDR)
lines(x = lambda, y = FDP, col = "black", lty = 2)  # dashed black line for FDP
lines(x = lambda, y = FDR, col = "black", lty = 1)  # solid black line for FDR
```
